---
title: "Shooting Deaths Project"
author: "S. Shasteen"
date: "2025-03-19"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
R.version
```


```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Read In Data & Check Column Names

```{r}
# Assign data URL to a variable 
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# Read in data
shooting_incidents <- read.csv(url)

# Check column names to get a sense of data

glimpse(shooting_incidents)

colnames(shooting_incidents)

```
## Fix Date/Time Data Types & Combine 2 Columns

```{r}
#Combine OCCUR_DATE and OCCUR_TIME into a single column

shooting_incidents <- shooting_incidents %>%
  mutate(OCCUR_DATETIME = mdy_hms(paste(OCCUR_DATE, OCCUR_TIME)))

# Verify new column

glimpse(shooting_incidents %>% select(OCCUR_DATE, OCCUR_TIME, OCCUR_DATETIME))

# Check for parsing failures

shooting_incidents %>%
  filter(is.na(OCCUR_DATETIME)) %>%
  select(OCCUR_DATE, OCCUR_TIME) %>%
  summarize(n_failed = n())

```
```{r}
# Check for NA/missing data.

sum(is.na(shooting_incidents)) #196 NA values total

# Check individual columns for NA/missing data.

colSums(is.na(shooting_incidents)) # Jurisdiction Code = 2; Lat = 97; Long = 97. = All 196. Inconsequential as I don't
                                   # care about lat/long or jurisdiction code. 


```
## Start Looking at Gender


```{r}

# Checking counts based on gender
shooting_incidents %>%
  count(VIC_SEX)

# Checking for missing or unique values in VIC_SEX

shooting_incidents %>%
  summarize(
    na_count = sum(is.na(VIC_SEX)),
    empty_count = sum(VIC_SEX == "" | VIC_SEX == "(null)", na.rm = TRUE),
    total_rows = n()
  )

# Checking for unique values

unique(shooting_incidents$VIC_SEX)

#Tally all values

shooting_incidents %>%
  count(VIC_SEX, sort = TRUE, .drop = FALSE) %>%
  mutate(percent = n/sum(n)*100)

```

## Subset the Data, Focus on Columns of Interest


```{r}
# Subset data to relevant columns

shoot_subset <- shooting_incidents %>%
  select(OCCUR_DATETIME, VIC_SEX, VIC_AGE_GROUP, PERP_SEX, PERP_AGE_GROUP, BORO, LOC_CLASSFCTN_DESC, LOCATION_DESC, STATISTICAL_MURDER_FLAG)

# Summarizing the subset

View(shoot_subset)





```

## Standardize Missing Values Under PERP_SEX

```{r}
# Standardize null or missing values to "U"

shoot_subset <- shoot_subset %>%
  mutate(PERP_SEX = case_when(
    PERP_SEX %in% c("M") ~"M",
    PERP_SEX %in% c("F") ~"F",
    PERP_SEX %in% c("U", "", "(null)") ~ "U",
    TRUE ~ NA_character_ 
  ))

# Verify standardization

shoot_subset %>%
  count(PERP_SEX, sort = TRUE) %>%
  mutate(percent = n / sum(n)*100)


```

## Look at Most Common Locations for Shootings

```{r}

# Subsetting the data

shoot_subset %>%
  count(PERP_SEX, sort = TRUE)

shoot_subset %>%
  count(LOCATION_DESC, sort = TRUE)

shoot_subset %>%
  count(LOC_CLASSFCTN_DESC, sort = TRUE)


```

## Standardize Missing Location Data

```{r}
# Handling missing data for locations

shoot_subset <- shoot_subset %>% 
  mutate(
    LOCATION_DESC = case_when(
      LOCATION_DESC %in% c("", "(null)") ~ "Unknown",
      TRUE ~ LOCATION_DESC
    ),
    LOC_CLASSFCTN_DESC = case_when(
      LOC_CLASSFCTN_DESC %in% c("", "(null)") ~ "Unknown",
      TRUE ~ LOC_CLASSFCTN_DESC
    )
  )


```

## Filter Male Killed by Female Incidents

```{r}
# Filter and count locations

male_killed_by_female <- shoot_subset %>% 
  filter(
    VIC_SEX == "M",
    PERP_SEX == "F",
      )

# Count LOCATION_DESC
male_killed_by_female %>%
  filter(LOCATION_DESC != "Unknown") %>%
  count(LOCATION_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

# Count LOC_CLASSFCTN_DESC
male_killed_by_female %>%
  filter(LOC_CLASSFCTN_DESC != "Unknown") %>%
  count(LOC_CLASSFCTN_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)


```

## Filter Female Killed by Male Incidents

```{r}

# Filter and count locations

female_killed_by_male <- shoot_subset %>% 
  filter(
    VIC_SEX == "F",
    PERP_SEX == "M",
      )

# Count LOCATION_DESC
female_killed_by_male %>%
  filter(LOCATION_DESC != "Unknown") %>%
  count(LOCATION_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

# Count LOC_CLASSFCTN_DESC
female_killed_by_male %>%
  filter(LOC_CLASSFCTN_DESC != "Unknown") %>%
  count(LOC_CLASSFCTN_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

```

## Filter Male Killed by Male Incidents

```{r}

# Filter and count locations

male_killed_by_male <- shoot_subset %>% 
  filter(
    VIC_SEX == "M",
    PERP_SEX == "M",
      )

# Count LOCATION_DESC
male_killed_by_male %>%
  filter(LOCATION_DESC != "Unknown") %>%
  count(LOCATION_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

# Count LOC_CLASSFCTN_DESC
male_killed_by_male %>%
  filter(LOC_CLASSFCTN_DESC != "Unknown") %>%
  count(LOC_CLASSFCTN_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

```

## Female Killed by Female Incidents

```{r}

female_killed_by_female <- shoot_subset %>% 
  filter(
    VIC_SEX == "F",
    PERP_SEX == "F",
      )

# Count LOCATION_DESC
female_killed_by_female %>%
  filter(LOCATION_DESC != "Unknown") %>%
  count(LOCATION_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

# Count LOC_CLASSFCTN_DESC
female_killed_by_female %>%
  filter(LOC_CLASSFCTN_DESC != "Unknown") %>%
  count(LOC_CLASSFCTN_DESC, sort = TRUE) %>% 
  mutate(percent = n / sum(n) * 100)

```



```{r}
# Saving this spot for what to do next


```

## Visualizing the Data

```{r}
# Combine gender pairs
combined <- bind_rows(
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "M") %>% 
    mutate(Group = "Women Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "F") %>% 
    mutate(Group = "Men Shot by Women"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "M") %>% 
    mutate(Group = "Men Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "F") %>% 
    mutate(Group = "Women Shot by Women")
) %>% 
  filter(LOCATION_DESC != "Unknown") %>% 
  count(Group, LOCATION_DESC) %>% 
  group_by(Group) %>% 
  mutate(percent = n / sum(n) * 100) %>% 
  # Keep top 5 locations per group
  slice_max(order_by = n, n = 5)

# Bar plot
ggplot(combined, aes(x = reorder(LOCATION_DESC, -percent), y = percent, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Top Shooting Locations by Victim-Perpetrator Gender",
    subtitle = "Percent of Non-Unknown Locations",
    x = "Location", y = "Percent", fill = "Gender Pair"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))


```


## Further Visualizations


```{r}
# Combine public housing shootings
public_housing <- bind_rows(
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "M", LOCATION_DESC == "MULTI DWELL - PUBLIC HOUS") %>% 
    mutate(Group = "Women Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "F", LOCATION_DESC == "MULTI DWELL - PUBLIC HOUS") %>% 
    mutate(Group = "Men Shot by Women"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "M", LOCATION_DESC == "MULTI DWELL - PUBLIC HOUS") %>% 
    mutate(Group = "Men Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "F", LOCATION_DESC == "MULTI DWELL - PUBLIC HOUS") %>% 
    mutate(Group = "Women Shot by Women")
) %>% 
  filter(!is.na(BORO), BORO != "") %>% 
  count(Group, BORO) %>% 
  group_by(Group) %>% 
  mutate(percent = n / sum(n) * 100)

# Bar plot
ggplot(public_housing, aes(x = BORO, y = percent, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Public Housing Shootings by Borough and Victim-Perpetrator Gender",
    subtitle = "Percent of Public Housing Shootings per Gender Pair",
    x = "Borough", y = "Percent", fill = "Gender Pair"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("Women Shot by Men" = "red", "Men Shot by Women" = "blue", 
                               "Men Shot by Men" = "green", "Women Shot by Women" = "purple"))
```


## Anothe Visualization. I'll Come Up with a Title Later


```{r}

# Combine gender pairs
combined_classfctn <- bind_rows(
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "M") %>% 
    mutate(Group = "Women Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "F") %>% 
    mutate(Group = "Men Shot by Women"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "M") %>% 
    mutate(Group = "Men Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "F") %>% 
    mutate(Group = "Women Shot by Women")
) %>% 
  filter(LOC_CLASSFCTN_DESC != "Unknown") %>% 
  count(Group, LOC_CLASSFCTN_DESC) %>% 
  group_by(Group) %>% 
  mutate(percent = n / sum(n) * 100)

# Bar plot
ggplot(combined_classfctn, aes(x = reorder(LOC_CLASSFCTN_DESC, -percent), y = percent, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Shooting Location Types by Victim-Perpetrator Gender",
    subtitle = "Percent of Non-Unknown Location Classifications",
    x = "Location Type", y = "Percent", fill = "Gender Pair"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))


```


## Domestic Violence Maybe? 

```{r}

# Perpetrator age for home settings
age_analysis <- bind_rows(
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "M", 
           LOCATION_DESC %in% c("MULTI DWELL - APT BUILD", "MULTI DWELL - PUBLIC HOUS", "PVT HOUSE")) %>% 
    mutate(Group = "Women Shot by Men"),
  shoot_subset %>% 
    filter(VIC_SEX == "M", PERP_SEX == "F", 
           LOCATION_DESC %in% c("MULTI DWELL - APT BUILD", "MULTI DWELL - PUBLIC HOUS", "PVT HOUSE")) %>% 
    mutate(Group = "Men Shot by Women"),
  shoot_subset %>% 
    filter(VIC_SEX == "F", PERP_SEX == "F", 
           LOCATION_DESC %in% c("MULTI DWELL - APT BUILD", "MULTI DWELL - PUBLIC HOUS", "PVT HOUSE")) %>% 
    mutate(Group = "Women Shot by Women")
) %>% 
  filter(PERP_AGE_GROUP != "UNKNOWN") %>% 
  count(Group, PERP_AGE_GROUP) %>% 
  group_by(Group) %>% 
  mutate(percent = n / sum(n) * 100)

# Bar plot
ggplot(age_analysis, aes(x = PERP_AGE_GROUP, y = percent, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Perpetrator Age in Home Setting Shootings",
    subtitle = "Percent of Non-Unknown Age Groups",
    x = "Perpetrator Age Group", y = "Percent", fill = "Gender Pair"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

```


# Conclusion

This analysis of NYPD shooting incidents (2006–2023, 29,744 cases) examines location patterns by victim and perpetrator gender, revealing distinct trends in gender-based violence:

- **Women Shot by Men** (1,830 cases): 82.4% of non-“Unknown” locations are home settings (35.6% apartment buildings, 32.9% public housing, 13.9% private houses), strongly indicating domestic violence, likely intimate partner violence in low-income settings (e.g., 25% Black poverty in Bronx).
- **Men Shot by Women** (380 cases): 84.6% of non-“Unknown” locations are home settings (39.9% apartment buildings, 30.3% public housing, 14.4% private houses), suggesting self-defense by women against male partners or, less commonly, female-on-male domestic violence.
- **Women Shot by Women** (small sample): 87.5% home settings (47.5% apartment buildings, 40.0% public housing), potentially reflecting domestic or interpersonal conflicts.
- **Men Shot by Men** (large sample): Higher public settings (7.5% grocery/bodega, 65.8% “STREET” in `LOC_CLASSFCTN_DESC`), consistent with community or gang-related violence (April 18 hypothesis).
- **Public Housing**: Prominent across groups (30.3–40.0% non-“Unknown”), possibly due to socioeconomic stress or policing bias (89% non-White stops in Brooklyn).

**Limitations**:
- High missing data: ~50% “Unknown” locations (46.9% women shot by men, 50.5% men shot by women), 41.8% unknown perpetrators, limiting pattern clarity.
- Small sample for men shot by women (380 cases) and women shot by women reduces precision.
- `LOC_CLASSFCTN_DESC` shows high “STREET” (e.g., 72% men shot by women), conflicting with `LOCATION_DESC` home dominance—possible data entry issues.
- No motive or legal outcome data to confirm self-defense vs. abuse.
- Excluded `STATISTICAL_MURDER_FLAG` due to unclear designation, combining fatal and non-fatal shootings.

**Future Steps**:
- Analyze `PERP_AGE_GROUP` to distinguish self-defense (25–44 females) vs. abuse (18–24 males).
- Cross-reference domestic violence reports for motive insights.
- Investigate `LOC_CLASSFCTN_DESC` inconsistencies (e.g., “STREET” vs. `LOCATION_DESC`).

**Key Statistics**:

```{r summary-table, echo=FALSE}
tibble(
  Group = c("Women Shot by Men", "Men Shot by Women", "Women Shot by Women", "Men Shot by Men"),
  Total_Cases = c(1830, 380, NA, NA),
  Home_Settings_Percent = c(82.4, 84.6, 87.5, NA),
  Public_Housing_Percent = c(32.9, 30.3, 40.0, 38.5)
) %>% 
  knitr::kable(caption = "Summary of Shooting Locations by Gender Pair")
```
